---
title: "Principal Component Analysis"
author: "Christopher J. Fiscus"
date: "updated 1/22/2018"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
## What is Principal Component Analysis (PCA)?
PCA a clustering method to reduce dimensionality in datasets and to explore patterns in data. The details of how this works can be found elsewhere.  


## Let's use the iris dataset
This dataset includes sepal length, sepal width, petal length, and petal width for 150 iris plants from three different species. 

```{r readdata, echo=TRUE}
## read in data
library(datasets)
df = iris

## Format of data (Name, lat, long)
head(df)
```

## Transform data
When doing PCA it is a good idea to transform the data so it is normalized. It's common to use a log transformation to do this. 

```{r transform, echo=TRUE, warnings=FALSE}
## log transformation on numerical data  
normalized<-log(iris[,1:4])
species<-iris[,5]

```

## Do PCA 
```{r pca, echo=TRUE, warnings=F}
## pca using single value decomposition, centered and scaled
df.pca<-prcomp(normalized, scale.=TRUE, center=TRUE)

```

## Plot the Percent Variance Explained by each PCA 
1. Format the data 
```{r pve, echo=TRUE, warnings=F}
## calculate percent variance
per.var<- (df.pca$sdev^2/sum(df.pca$sdev^2))*100
per.var<-as.data.frame(per.var)

## format data nicely 
per.var<-cbind(1:nrow(per.var), per.var)
colnames(per.var)<-c("PC", "PercentVariance")

```
2. plot with ggplot2 
```{r plot_pervar, echo=TRUE, warnings=F}
library(ggplot2)
g<-ggplot(data=per.var, aes(x=factor(PC), y=PercentVariance, axis.ticks=1)) + geom_bar(stat="identity") + xlab("principal component") + ylab("% Variance") + theme(text=element_text(size=14)) + ggtitle("Variance Explained by PCs") + theme_classic()
print(g)
# write graph out 
# ggsave("varexplained.png", plot=g)

```

## Plot PCs 1-3 
1. Plot PC 1 vs. PC 2
Note: The github version of ggfortify will add the % variance explained to the axis. The CRAN version does not have this yet. 
```{r plot_pca, echo=TRUE, warnings=F}
library(factoextra)

## install github version of ggfortify 
#library(devtools)
#install_github('sinhrks/ggfortify')

library(ggfortify) # load lib

## Plot PC1 vs. PC2
g<-autoplot(df.pca, data=df, colour='Species', alpha=0.5, variance_percentage=TRUE, loadings=FALSE, scale=0) + theme_classic()
print(g)
```
2. Plot PC2 vs. PC 3
```{r plot_pca2, echo=TRUE, warnings=F}
## Plot PC2 vs. PC3
g<-autoplot(df.pca, data=df, colour='Species', alpha=0.5, variance_percentage=TRUE, loadings=FALSE, x=2, y=3, scale=0) + theme_classic()
print(g)
```

## Explore loadings
```{r loadings, echo=TRUE, warnings=F}
## Loadings for PCs 1-3
Loadings<-as.data.frame(df.pca$rotation[,1:3])
head(Loadings)
```

## Plot Densities of PCs 1 - 3 
```{r density, echo=TRUE, warnings=F}
g<-ggplot(Loadings) + geom_density(aes(PC1, color="black")) + theme_classic() + geom_density(aes(PC2, color="red")) + geom_density(aes(PC3, color="blue")) + xlab("Contribution to PC") + theme(text=element_text(size=16)) + scale_color_manual(labels=c("1", "2", "3"), values=c("black", "red", "blue")) + labs(colour="PC")
print(g)
```

## Explore PCA
```{r contrib, echo=TRUE, warnings=F}
res.ind <- get_pca_ind(df.pca)

## coordinates
coord<-as.data.frame(res.ind$coord)
head(coord)

## contributions 
contri<-as.data.frame(res.ind$contrib)
head(contri)

## quality
qual<-as.data.frame(res.ind$cos2)
head(qual)
```
